### 2024-05-23 thesis/ms-fletcher-porter.tex
Integrate the JFPS work into the thesis

%% Use this if you want to get links and nice output. Works well with pdflatex.
\usepackage{hyperref}
\hypersetup{
  pdfpagemode=UseNone,
  pdfstartview=FitH,
  colorlinks=true,
  urlcolor=xkcd_blue,
  linkcolor=black,
  citecolor=black,
  pdftitle=A Model for Selecting Software Tools for Fluid Power Systems,
  pdfauthor=Fletcher Porter,
  pdfkeywords={
    Mechatronics,Tooling,Automation,Instrumentation,Tech Stack
  }
}

@@

%% English abstract.
%% All the information required in the abstract (your name, thesis title, etc.)
%% is used as specified above.
%% Specify keywords
%%
%% Kaikki tiivistelmässä tarvittava tieto (nimesi, työnnimi, jne.) käytetään
%% niin kuin se on yllä määritelty.
%% Avainsanat
%%
\keywords{Mechatronics,Tooling,Automation,Instrumentation,Tech Stack}
%% Abstract text
\begin{abstractpage}[english]
Mechatronic systems can be enormously complex.  Possibly the chiefest complexity is in their software components due the the deep knowledge required for effective development, the need to interface with myriad heterogeneous components, and the opacity of its operation.  I'll be speaking about alleviating these issues by introducing a novel theory for specifying and discussing the need for software tools and then a case study in applying this theory in the development of a Stewart Platform used as a driving simulator platform. 
\end{abstractpage}

@@

%% Symbols and abbreviations
\mysection{Symbols and abbreviations}

\subsection*{Symbols}

\begin{tabular}{ll}
$\mathbf{B}$  & magnetic flux density  \\
$c$              & speed of light in vacuum $\approx 3\times10^8$ [m/s]\\
$\omega_{\mathrm{D}}$    & Debye frequency \\
$\omega_{\mathrm{latt}}$ & average phonon frequency of lattice \\
$\uparrow$       & electron spin direction up\\
$\downarrow$     & electron spin direction down
\end{tabular}

\subsection*{Operators}

\begin{tabular}{ll}
$\nabla \times \mathbf{A}$              & curl of vectorin $\mathbf{A}$\\
$\displaystyle\frac{\mbox{d}}{\mbox{d} t}$ & derivative with respect to 
variable $t$\\[3mm]
$\displaystyle\frac{\partial}{\partial t}$  & partial derivative with respect 
to variable $t$ \\[3mm]
$\sum_i $                       & sum over index $i$\\
$\mathbf{A} \cdot \mathbf{B}$    & dot product of vectors $\mathbf{A}$ and 
$\mathbf{B}$
\end{tabular}

\subsection*{Abbreviations}

\begin{tabular}{ll}
  IT & Information Technology \\
  HTML & Hyper Text Markup Language \\
  SQL & Structured Query Language \\
  CAN & Controller Area Network \\
  OSI & Open Systems Interconnection \\
  DDH & Directy-Drive Hydraulics \\
  EHA & Electro-Hydraulic Actuator \\
  ECU & Electronic Control Unit \\
  DAQ & Data Acquisition \\
  IO & Input/Output \\
  PLC & Programmable Logic Controller \\
  TCP & Transmission Control Protocol \\
  IP & Internet Protocol \\
  UDP & User Datagram Protocol \\
  IDE & Integrated Development Environment \\
\end{tabular}

@@

Research into fluid power systems is difficult and complex.  When doing research with hardware systems, before a single research question can be asked of the hardware, researchers must assemble mechanical components, hook up power electronics, and route hoses for the hydraulic and pneumatic systems.  Then the most complex of them all, the control system must be added.  Its tangle of spindly wires reaches all throughout the machine, meeting at the computer, the single most complex component of any fluid power system.

This paper improves the process of commissioning control systems for fluid power systems.  It will do this by reasoning about how to abstract behavior and components in control systems into layers that are more easily reasoned with, synthesizing a novel {\it mechatronic tech stack theory}.  It will go on to demonstrate in three case studies how this theory can be applied in fluid power systems.

@@

This section constructs the backbone of this thesis, namely the novel theory of a mechatronic tech stack.  First, there will be a review of tech stacks and {\it abstraction layers\/} as they've been applied in the IT sector.  Then there will be a proposal of a mechatronic tech stack determined by a standards-based set of principals.

@@

The concept of a tech stack, alternately called a solution stack or a software stack, is to enumerate the tasks done by a kind of software application and to suggest tools for those tasks~\cite{PranamStack2017}.  More specifically, it defines a set of abstraction layers; each layer abstractly describes a type of task and contains within it the set of software tools that complete that task.  The intention is to take the difficult task of tool selection and provide a structure through which practitioners can think about the functions and tooling needs of their application, freeing mental energy to critically assess which tools would be best for their technical and personnel needs.

@@

A ubiquitous example of a tech stack in IT is the {\it web stack}, used for developing websites and related pieces of software.  The tasks that websites must perform, as suggested by this stack, are to present an interface to a user through a web browser (the {\it front end\/}), to conduct some business logic like creating a customized page for the user (the {\it back end\/}), to perform network communication (the {\it server\/}), and to store and dispatch data (the {\it database\/})~\cite{PranamStack2017}.  This doesn't describe all websites---many don't have databases or any real back-end logic, but it describes at least a superset of the functionality of most websites.

Tools in each of these layers are myriad and to highlight that, proceeding is a list of a small sample.  You can also observe this in figure~\ref{fig:web_stack}  Front-end tools like jQuery~\cite{jQuery} have been proven from many years of use, React~\cite{React} is extraordinarily powerful and can be used in non-web contexts as well, and even plain HTML without an additional framework can be good on account of its simplicity.  Backend tools can allow a developer to select their favorite programming language, or one that is specialized for web backends, or even JavaScript like the front end uses.  Servers are often integrated into the back-end framework, but if not, specialized servers like Apache~\cite{ApacheServer} or nginx~\cite{nginx} are popular.  Databases are dominated by reliable and proven relational, SQL-based offerings, but alternatives like the confusingly named NoSQL or Redis~\cite{redis} have become popular in the past years on account of their performance and scalability.

\begin{figure}[h]
  \centering
  \includegraphics[width=0.5\textwidth]{assets/web_stack}
  \caption{The web stack is like four bins of tools, here illustrated as boxes with the logos of common tools within them.}\label{fig:web_stack}
\end{figure}

In discussing the web stack, it's also useful to think about a particular set of tools used for a single project, i.e.\ one tool from each layer.  Alternately, this can be thought of as the project's {\it cross section\/} of the web stack, or the the project's own stack.  There are several of these cross sections that are widely used, such as LAMP (Linux, Apache, MySQL, PHP) or MEAN (MongoDB, Express.js, Angular.js, Node.js)~\cite{PranamStack2017}.  Just as often, firms employ stacks tailored to their technical and personnel needs and often write about these decisions in blog posts, such as for recent changes to Microsoft Teams in~\cite{Singh2023}.

@@

Another example of a tech stack is the OSI Network model which abstractly describes network communication down to the lowest levels of physical signaling up to the highest levels of abstract behaviour~\cite{ISO7498-1}.  This model is frequently cited in mechatronic literature since many fieldbus protocols like CAN define their functionality in terms of this model.  This model is also interesting since its standard description (ISO~7498-1) also includes principal for creating layered models of a similar ilk.

The OSI model differs from the web stack in how the layers are used by practitioners.  Typically, instead of each layer being a bin to pick a tool from and where a tools must be picked from every layer which you want functionality, developers instead may decide at what depth into the model they begin their implementation and assume everything below that depth ``just works''.

To spare a moment for the details of the model, it consists of layered abstraction over physical communication media, as seen in figure~\ref{fig:osi_model}.  The lowest layer, Physical, concerns itself with the physical phenomena of communication.  In the case of CAN, this means voltage levels, terminating resistors, twisted pairs, grounding rules, and more.  The highest level, Application, provides network facilities without need for understanding {\it how\/} the network connection is made.  The Application layer includes, for example, identifying communication partners and negotiating security protocols.  Well-known protocols like HTTP, DNS, and FTP exist in the Application layer.  The intermediate layers basically provide increasing levels of abstraction over the physical layer, defining things like the structure of data packets, how to route messages, how to route them efficiently, and more.

\begin{figure}[h]
  \centering
  \includegraphics[width=0.6\textwidth]{assets/osi_model}
  \caption{The OSI model understands network communication to be a layer cake of abstractions over the physical communication media.  Taken from~\cite[§6.1.3]{ISO7498-1}.}\label{fig:osi_model}
\end{figure}

An interesting property of this layered abstraction architecture is that protocols which are defined in higher layers can often have the lower layers transparently replaced.  For instance, the internet protocol which we use every day is defined in the Network and Transport layers.  Typically the lower-level protocols used with IP are Ethernet and IEEE~802.11 (Wi-Fi).  However, with some implementation effort, other low-level protocols are available.  Lindgren et al.\ developed IP over CAN, wherein an internet network is extended over a CAN network~\cite{Lindgren2008}, something which was never imagined by implementers of CAN nor IP.\

@@

The OSI standard also includes the principals upon which its layers were designed~\cite[§6.2.1]{ISO7498-1}.  In summary, each layer should be a discrete, separable item with a coherent function.  The interfaces between layers should be succinct and only interface with adjacent layers.  It suggests having fewer layers would ease explanation and implementation and it offers a test to aid in determining layer boundaries.  It also says that the layers should be motivated by experience.  Being a novel theory, experience will be the principal gain of future work, although case studies discussed later will provide somewhat in this dimension.  A layer should have a different level of abstraction in terms of {\it morphology}, {\it syntax}, and {\it semantics}.

This model of morphology, syntax, and semantics come from Bloomfieldian linguistics~\cite{Bloomfield1923}.  It's used to coerce natural languages into a structured form from which meaning can be gleaned.  Here, the standard authors are using it to glean structure and meaning from engineering systems.  To understand this model, take this line of C code.

@@

Morphology is concerned with the atomic elements of meaning.  In C code, this means short tokens of text (also called morphemes), here \verb|printf|, \verb|(|, \verb|"I have %u cows\n"|, \verb|,|, \verb|num_cows|, \verb|)|, and \verb|;|.  None of these can be cut in half and still carry meaning---\verb|pri| and \verb|ntf| doesn't make \verb|printf|.  But these atoms of meaning needn't just be text or language.  As will be discussed below, these atoms can be cables, connectors, computer hardware, or really anything that can have meaning when combined with other things.

When these morphemes are combined into phrases, that's syntax.  Syntax is the structure into which meaning is applied.  The \verb|printf| example is syntactically parsed by C compilers as in figure~\ref{fig:printf_syntax}.  The syntax tells that the postfix expression represented by the identifier \verb|printf| has the argument list of \verb|"I have %u cows\n"| (a string literal) and \verb|num_cows| (an identifier).  What any of these tokens mean is not the domain of syntax, but instead semantics.

@@

To understand the semantics of \verb|printf|, you must read the documentation.  It will tell you that \verb|printf| means to convert and write output to the console under the control of a format string.  You'll also see that \verb|"I have %u cows\n"| is that format string, and \verb|num_cows| is an additional argument, presumably the number of cows the author has.  In fact, reading the documentation further, you'll learn that the format string itself has its own morphology, syntax, and semantics driving how formatted output is created.

@@

In the next section, this knowledge will be applied to create a mechatronic tech stack with the properties of the web and OSI stacks.

@@

Broadly speaking, mechatronic systems include mechanical systems, power-electronic systems, and digital control systems (neglecting mechanical governors and analog control circuits).  All of these could be considered under a theory of a mechatronic tech stack, but the present treatment will only consider the control systems.  More specifically, this stack will start just after any sensors or transducers, giving those items IO (input/output) hookups.

The layers will by identified by noting contrasts in morphology, syntax, and semantics throughout the gamut of the control system.  Layer boundaries will become evident where differences in these aspects are dramatic.  I'll then describe the layers and the interfaces between them.

Morphology provides the strongest and most coarse distinctions in layers.  If the atoms that one component is made of are incompatible with another, they can't effectively be abstracted together.  The most obvious case is that those components that are made of literal atoms, i.e.\ hardware, must be considered separately from those that aren't physical, namely software.  Within hardware, a control engineer will concern themselves with the components of the computer which their control softwares run on (RAM, CPU, storage, etc.) and the physical connections between the computer components and the sensors and transducers, such as cables, connectors, contacts, and traces.  Making up the software is the instructions of the control program itself, but also supporting software like the operating system, runtime environments, and useful libraries.

Syntax is the structure of an idea.  It is a finer comb to discriminate layers.  Software generally is built as a linear sequence of instructions with branches and loops.  This describes the control engineer's perspective on their own control program, but supporting software like libraries are more abstract.  They're better thought of as a dependency.  A control program may depend on libraries A, B, and C, and those in turn depend on libraries D, E, F, and G.  These {\it middlewares\/} are better understood as a network of the softwares themselves and the dependency connections between them.  For hardware, its structure is its assemblage.  The computer components are connected to the motherboard, sensors and transducers are connected to a CAN bus, the temperature sensors communicate over I2C, etc.  Like middleware, hardware is the network of components and the connections between them, but the type of connection is also important.  A system with components on a CAN bus is going to be different than one with a USB bus, and different still from a system where all components communicate over a serial connection.

Semantic meaning reveals the most subtle distinctions in layers.  The control program is an expression of the business logic desired by the control engineer.  It is whatever control logic that is asked of the engineer.  Middleware begins where the controller ends, which is to say it is abstraction around the computer's function.  It attempts to save the engineer from difficult, tedious, or uninteresting implementation.  The middleware ends where the computer begins.  Computers are meaningful when they provide compelling output based on some input.  That is to say, when they compute.  How this computation occurs at the hardware level is determined by the computers {\it architecture}.  This defines what machine instructions do what, which hardware pins go where, and how it should all be connected.  IO is the hardware beyond the computer.  Its is the language of data transfer.  A sensor's pressure reading becomes a message to the computer.  The computer's actuation command becomes a message to the transducer.  This is handled by network communication protocols, the domain of the OSI model.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=0.5\textwidth]{assets/mechatronic_tech_stack}
  \caption{The mechatronic tech stack places high-level business logic at the top and low-level interfacing at the bottom.  This sits atop hardware instrumentation, where this theory placed its limit.}\label{fig:mechatronic_tech_stack}
\end{figure}

Between the layers are the interfaces that enable higher layers to take advantage to take advantage of the abstractions provided by lower layers.  The control layer, being the highest, provides no interface to further layers, but from the middleware layer it demands certain {\it application programming interfaces}, or APIs.  Each program in the middleware layer (generally) provides an API, so the control-middleware interface is the union of all of these APIs.  The needs of middleware are typically met by other middleware, but those that interface with the computer directly do so through architecture-defined machine code.  Then for the Computer-IO interface, the IO needs some hardware connection like pins broken out from the CPU, integral connection ports like for USB, serial, or RJ45 (Ethernet).  Finally, IO interfaces with sensors and transducers in the same way it does with computers, through hardware interfaces like terminal blocks or hookups to a fieldbus.

This model, like the web stack, provides a structure through which to think about tool selection.  The choice of using CAN- or serial-based IO can be made separately from the decision of what programming language to use for the controller.  This isn't to say that tooling decision in the high layers is completely decoupled from those in the low layers.  The decision to use an AVR-based Arduino computer likely precludes writing the controller in JavaScript, but this is due to the middleware to support that being unavailable or impossible.

Following will be several case studies providing concrete illustrations of how this mechatronic tech stack model can be applied.  These will mostly be descriptive case studies that apply this theory for retrospective research, but one case uses this model prescriptively.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=0.4\textwidth]{assets/dolores}
  \caption{Dolores rotates the top platform about a pivot by actuating a pair of hydraulic cylinders.  A pair of additional cylinders in parallel can resist the motion providing loads useful for testing}\label{fig:dolores0}
\end{figure}

The industry partner's requested change occurred in the IO layer.  The old system has an inverter that drives a servomotor, a safety valve to enable the system, and an array of pressure, temperature, and flow sensors to observe the state of the hydraulic system.  The inverter and the pressure and temperature sensors all communicated over the CANopen protocol, the flow sensors gave analog inputs, and the safety valved needed a digital output.  The CAN communication was handled by a Vector CAN/USB dongle~\cite{VectorCanUsbDongle}.  All other IO went through a National Instruments data acquisition device (DAQ).  In the new system, the forced change was that the inverter had to communicate over EtherCAT.\  This meant there needed to be new hardware to support this protocol.  That hardware was selected to be Beckhoff EtherCAT terminals.  They provided the interface to connect the inverter to the computer and also to provide digital and analog IO as well as a CAN bus to communicate with the extant sensor network~\cite{BeckhoffEL6751}.  This made the DAQ unnecessary, but the CAN/USB dongle was still required due to deficiencies in Beckhoff's control interfaces which are discussed below.

The system had been controlled by a conventional Windows workstation.  The DAQ had connected through a PCIe slot on the workstation's motherboard.  This may have been fine to keep for the new system, but it would have tied the workstation down to Dolores instead of being used for the several experiments in the lab.  One of Beckhoff's industrial PCs was selected as the computer in order to leverage Beckhoff's support~\cite{BeckhoffC60XX}.  It was fairly expensive, but it was thought that architecting the system to align with vendor expectations would make development easier and cause fewer problems to arise.  However, the workstation did tap into the CAN bus to initialize the sensors, so it wasn't completely out of the loop.

In the old system, the workstation ran Windows, the DAQ was managed by drivers from National Instruments, and the CAN bus was initialized using Vector's CANalyzer application and runtime commands came through Vector's Simulink library~\cite{VectorCanalyzer, SimulinkVectorCAN}.  Due to issues getting reliable connections over EtherCAT with Simulink (possibly due to user error, though the documentation was hardly simple), the new system uses a PLC-based middleware.  Still based on Windows, it has libraries and runtimes to run the EtherCAT bus, to read from the CAN bus, and to interface with an external development computer.  CANalyzer remained to initialize the CAN bus.

The old system had controllers written with Matlab and Simulink.  However, due to middleware issues, they were abandoned in favor of writing controllers in PLC languages based on IEC~61131-3 with the TwinCAT development environment~\cite{IEC61131-3, BeckhoffTwinCAT}.  The major issue with TwinCAT is that while it nominally supported sending messages over CANopen, there was no documented way to initialize CAN devices.  This was fixed by manually sending commands with CANalyzer at initialization time.

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{assets/mechatronic_tech_stack_dolores}
  \caption{The change was to use Beckhoff's stack, though it didn't work without supplements.}\label{fig:mechatronic_tech_stack_dolores}
\end{figure}

>>>>

%% Use this if you want to get links and nice output. Works well with pdflatex.
\usepackage{hyperref}
\hypersetup{
  pdfpagemode=UseNone,
  pdfstartview=FitH,
  colorlinks=true,
  urlcolor=xkcd_blue,
  linkcolor=black,
  citecolor=black,
  pdftitle=A Model for Selecting Software Tools for Fluid Power Systems,
  pdfauthor=Fletcher Porter,
  pdfkeywords={
    Mechatronics,Control Engineering,Software Tooling,Stewart Platform,Pneumatic actuator
  }
}

@@

%% English abstract.
%% All the information required in the abstract (your name, thesis title, etc.)
%% is used as specified above.
%% Specify keywords
%%
%% Kaikki tiivistelmässä tarvittava tieto (nimesi, työnnimi, jne.) käytetään
%% niin kuin se on yllä määritelty.
%% Avainsanat
%%
\keywords{Mechatronics,Control Engineering,Software Tooling,Stewart Platform,Pneumatic actuator}
%% Abstract text
\begin{abstractpage}[english]
Fluid power systems can be enormously complex. The chiefest complexity is in their control components due to the deep knowledge required for effective development, the need to interface with myriad heterogeneous components, and the opacity of its operation. In the IT sector, a tech stack is a known discretization of function within a system, typically used to categorize what software tools do. This enables disparate parts of a system to be developed separately. Inspired by IT tech stacks, this paper presents a model, the mechatronic tech stack, to mitigate this complexity by siloing control system behavior into discrete functions and classifying what tools can be used to meet them. The model is demonstrated on three case studies of diverse application. The theory is promising, but there are challenges in defining the most useful abstractions and interfaces.
\end{abstractpage}

@@

%% Symbols and abbreviations
\mysection{Abbreviations}

\begin{tabular}{ll}
  IT & Information Technology \\
  HTML & Hyper Text Markup Language \\
  SQL & Structured Query Language \\
  CAN & Controller Area Network \\
  ISO & International Organization for Standardization \\
  OSI & Open Systems Interconnection \\
  HTTP & Hypertext Transfer Protocol \\
  DNS & Domain Name System \\
  FTP & File Transfer Protocol \\
  IP & Internet Protocol \\
  IO & Input/Output \\
  RAM & Random Access Memory \\
  CPU & Central Processing Unit \\
  API & Application Programming Interface \\
  I2C & Inter-Integrated Circuit \\
  USB & Universal Serial Bus \\
  PLC & Programmable Logic Controller \\
  DAQ & Data Acquisition System \\
  IEC & International Electrotechnical Commission \\
  EFPA & Extension type Flexible Pneumatic Actuator \\
  EEPROM & Electronically-Erasable Programmable Read-Only Memory \\
\end{tabular}

@@

In order to study mechatronics systems, it is necessary to design and develop physical research testbeds. This is especially the case in mechatronic systems with fluid components, as their high-fidelity modelling is challenging and computationally expensive. When doing research with hardware systems, before a single research question can be asked of the hardware, researchers must assemble mechanical components, hook up power electronics, and route hoses for the hydraulic and pneumatic systems. Then the most complex of them all, the control system, must be added. Its tangle of spindly wires reaches all throughout the machine, meeting at the computer, the single most complex component of any fluid power system. The requirements placed on it are immense. In high-power systems, safety invariants must be held up perfectly lest disaster strike, and close monitoring is necessary due to the low stiffness of fluid systems compared to mechanical ones.

This paper improves the process of commissioning control systems for fluid power systems by reasoning about how to abstract behavior and components in control systems into layers, synthesizing a novel {\it mechatronic tech stack\/} theory. It will demonstrate in three case studies how this theory can be applied in fluid power systems. The tools used in these cases will be categorized into layers of the mechatronic tech stack.  The analysis shows that, while different, these cases all fit into the same structure.

@@

This section constructs the backbone of this paper, namely the novel theory of a mechatronic tech stack. First, there will be a review of tech stacks and {\it abstraction layers\/} as they've been applied in the IT sector. Then there will be a proposal of a mechatronic tech stack determined by a standards-based set of principles.

@@

The concept of a tech stack, alternately called a solution stack or a software stack, is to enumerate the tasks done by a kind of software application and to suggest tools for those tasks~\cite{PranamStack2017}. More specifically, it defines a set of abstraction layers; each layer describes a type of task and contains within it the set of software tools that complete that task. The intention is to take the difficult task of tool selection and provide a structure through which practitioners can think about the functions and tooling needs of their application, freeing mental energy to critically assess which tools would be best for their technical and personnel needs.

@@

A ubiquitous example of a tech stack in IT is the \textit{web stack}, used for developing websites and related pieces of software. The task that websites must perform, as suggested by this stack, is to present an interface to a user through a web browser (the \textit{front end}), to conduct some business logic like creating a customized page for the user (the \textit{back end}), to perform network communication (the \textit{server}), and to store and dispatch data (the \textit{database})~\cite{PranamStack2017}. This doesn't describe all websites---many don't have databases or any real back-end logic, but it describes at least a superset of the functionality of most websites.

Tools in each of these layers are myriad and to highlight that, proceeding in figure~\ref{fig:web_stack} is a list of a small sample. Front-end tools like jQuery~\cite{jQuery} have been proven from many years of use, React~\cite{React} is very powerful and can be used in non-web contexts, and even plain HTML without an additional framework can be good on account of its simplicity. Back-end tools can allow a developer to select their favorite programming languages like Ruby or Python, or one that is specialized for web back-ends like PHP, or even JavaScript like the front end uses. Servers are often integrated into the back-end framework, but if not, specialized servers like Apache~\cite{ApacheServer} or nginx~\cite{nginx} are popular. Databases are dominated by reliable and proven relational, SQL-based offerings like MySQL or PostgreSQL, but alternatives like NoSQL or Redis~\cite{redis} have become popular in the past years.

\begin{figure}[h]
  \centering
  \includegraphics[width=0.7\textwidth]{assets/web_stack}
  \caption{The web stack is like four bins of tools, here illustrated as boxes with the logos of common tools within them.}\label{fig:web_stack}
\end{figure}

In discussing the web stack, it's also useful to think about a particular set of tools used for a single project, i.e.\ one tool from each layer. Alternately, this can be thought of as the project's cross section of the web stack, or the project's own stack. There are several of these cross sections that are widely used, such as LAMP (Linux, Apache, MySQL, PHP) or MEAN (MongoDB, Express.js, Angular.js, Node.js)~\cite{PranamStack2017}. Just as often, firms employ stacks tailored to their technical and personnel needs and often write about these decisions in blog posts, such as for recent changes to Microsoft Teams in~\cite{Singh2023}.

@@

Another example of a tech stack is the OSI Network model which abstractly describes network communication down to the lowest levels of physical signaling up to the highest levels of abstract behavior~\cite{ISO7498-1}. This model is frequently cited in mechatronic literature since many fieldbus protocols like CAN define their functionality in terms of this model. This model is also interesting since its standard description (ISO~7498-1) also includes principles for creating layered models of a similar ilk.

The OSI model differs from the web stack in how the layers are used by practitioners. Typically, instead of each layer being a bin to pick a tool from and where a tool must be picked from every layer which you want functionality, developers instead may decide at what depth into the model they begin their implementation and assume everything below that depth ``just works''.

The model itself consists of layered abstractions over physical communication media, as seen in figure~\ref{fig:osi_model}. The lowest layer, Physical, concerns itself with the physical phenomena of communication. In the case of CAN, this means voltage levels, terminating resistors, twisted pairs, grounding rules, and more. The highest level, Application, provides network facilities without need for understanding \textit{how} the network connection is made. The Application layer includes, for example, identifying communication partners and negotiating security protocols. Well-known protocols like HTTP, DNS, and FTP exist in the Application layer. The intermediate layers provide increasing levels of abstraction over the physical layer, defining things like the structure of data packets, how to route messages, how to route them efficiently, and more.

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{assets/osi_model}
  \caption{The OSI model understands network communication to be a layer cake of abstractions over the physical communication media.  Taken from~\cite[§6.1.3]{ISO7498-1}.}\label{fig:osi_model}
\end{figure}

An interesting property of layered abstraction architectures is that protocols which are defined in higher layers can have the lower layers transparently replaced. For instance, the internet protocol which we use every day is defined in the Network and Transport layers. Typically, the internet is built over Ethernet or Wi-Fi, but with some effort, other low-level protocols are available. Lindgren et al.\ developed IP over CAN, wherein an internet network is extended over a CAN network~\cite{Lindgren2008}, which was never imagined by implementers of CAN nor IP.

@@

The OSI standard also includes the principles upon which its layers were designed~\cite[§6.2.1]{ISO7498-1}. In summary, each layer should be a discrete, separable item with a coherent function. Their interfaces should be succinct and only connect with adjacent layers. Fewer layers eases explanation and implementation and the choice layers should be motivated by experience. Being a novel theory, experience will be the principal gain of future work, although case studies discussed later will also help. Usefully, it offers a test to aid in determining layer boundaries. Layers should have a different level of abstraction in terms of \textit{morphology}, \textit{syntax}, and \textit{semantics}.

This model of morphology, syntax, and semantics comes from Bloomfieldian linguistics~\cite{Bloomfield1923}. It's used to coerce natural language into a structured form from which meaning can be gleaned. Here, the standard authors are using it to glean structure and meaning from engineering systems. To understand this theory, take this line of C code.

@@

Morphology is concerned with the atomic elements of meaning. In C code, this means short tokens of text, here \verb|printf|, \verb|(|, \verb|"I have %u cows\n"|, \verb|,|, \verb|num_cows|, \verb|)|, and \verb|;|. None of these can be reduced and still carry meaning---\verb|pri| and \verb|ntf| doesn't make \verb|printf|. But these atoms of meaning needn't be text or language. As will be discussed, they can be cables, connectors, computer hardware, or anything with relational meaning.

When these tokens are combined into phrases, that's syntax. Syntax is the structure into which meaning is applied. The \verb|printf| example is syntactically parsed by C compilers as in figure~\ref{fig:printf_syntax}. The syntax tells that the postfix expression represented by the identifier printf has the argument list of \verb|"I have %u cows\n"| (a string literal) and \verb|num_cows| (an identifier). What any of these tokens mean is the domain of semantics.

@@

To understand the semantics of \verb|printf|, you must read the documentation. It will tell you that \verb|printf| means to convert and write output to the console under the control of a format string. You'll also see that \verb|"I have %u cows\n"| is that format string, and \verb|num_cows| is an additional argument, presumably a number of cows. Reading the documentation further, you'll see that the format string itself has its own morphology, syntax, and semantics driving how output is created.

@@

Next, these ideas will be used to make a mechatronic tech stack like the web and OSI stacks.

@@

Broadly speaking, mechatronic systems include mechanical systems, power-electronic systems, and digital control systems. All of these could be considered under a theory of a mechatronic tech stack, but the present treatment will only consider the control systems. Specifically, the stack will start at the IO (input/output) interface for sensors and transducers.

The layers will be identified by noting contrasts in morphology, syntax, and semantics throughout the gamut of the control system. Layer boundaries will become evident where there are dramatic differences. I'll then describe the layers and the interfaces between them.

Morphology provides the strongest and most coarse distinctions in layers. If the atoms that one component is made of are incompatible with another, they can't effectively be considered together. The most obvious case is that those components that are made of literal atoms, i.e.\ hardware, must be considered separately from those that aren't physical, namely software. Within hardware, a control engineer will concern themselves with the components of the computer which their control software run on (RAM, CPU, storage, etc.) and the physical connections between the computer components and the sensors and transducers, such as cables, connectors, contacts, and traces. The software is the instructions of the control program itself, but also supporting software like the operating system, runtime environments, and libraries.

Syntax is the structure of an idea. It is a finer comb to discriminate layers. Software generally is built as a linear sequence of instructions with branches and loops. This describes the control engineer's perspective on their own control program, but supporting software like libraries are more abstract. They're better thought of as a dependency. A control program may depend on libraries A, B, and C, and those in turn depend on libraries D, E, F, and G. These \textit{middlewares} are better understood as a network of the programs themselves and the dependency connections between them. For hardware, its structure is its assemblage. The computer components are connected to the motherboard, sensors and transducers are connected to a CAN bus, the temperature sensors message over I2C, etc. Like middleware, hardware is the network of components and the connections between them, but the type of connection is also important. A system with components on a CAN bus is going to be different than one with a USB bus, and different still from a system where all components communicate over a serial connection.

Semantic meaning reveals the most subtle distinctions in layers. The control program is an expression of the business logic desired by the engineer. Middleware begins where the controller ends. It attempts to save the engineer from difficult, tedious, or uninteresting implementation otherwise necessary with computers. Computers are meaningful when they provide compelling output based on some input, i.e.\ when they compute. How this computation occurs at the hardware level is determined by the computer's \textit{architecture}. This defines what machine instructions do what and which hardware pins go where. IO is the hardware beyond the computer. It works on data transfer. A sensor's pressure reading becomes a message to the computer. The computer's actuation command becomes a message to the transducer. This is handled by e.g.\ network communication protocols, the domain of the OSI model.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=0.7\textwidth]{assets/mechatronic_tech_stack}
  \caption{The mechatronic tech stack places high-level business logic at the top and low-level interfacing at the bottom.}\label{fig:mechatronic_tech_stack}
\end{figure}

Between the layers are the interfaces that enable higher layers to take advantage of the abstractions provided by lower layers. The control layer, being the highest, provides no interface to further layers, but from the middleware layer it demands certain \textit{application programming interfaces}, or APIs. Each program in the middleware layer generally provides an API, so the control-middleware interface is the union of all these APIs. The needs of middleware are typically met by other middleware, but those that interface with the computer directly do so through architecture-defined machine code. Then for the Computer-IO interface, the IO needs hardware connections like pins broken out from the CPU or connection ports like for CAN, serial, or Ethernet. Finally, IO interfaces with sensors and transducers in the same way it does with computers, through hardware interfaces like terminal blocks or hookups to a fieldbus.

This model, like the web stack, provides a structure through which to think about tool selection. The choice of using CAN- or serial-based IO can be understood separately from the decision of what programming language to use for the controller. This isn't to say that tooling decision in the high layers is completely decoupled from those in the low layers. The decision to use an AVR-based Arduino computer likely precludes writing the controller in JavaScript, but this is due to a lack of middleware to support that use case.

Following will be three case studies providing concrete illustrations of how this mechatronic tech stack model can be applied. These will mostly be descriptive case studies that apply this theory for retrospective research, but one case uses this model prescriptively.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=0.6\textwidth]{assets/dolores}
  \caption{Dolores rotates the top platform about a pivot by actuating a pair of hydraulic cylinders.  A pair of additional cylinders in parallel can resist the motion providing loads useful for testing.}\label{fig:dolores0}
\end{figure}

The industry partner's requested change occurred in the IO (input/output) layer. The old system had an inverter that drives a servomotor, a safety valve to enable the system, and an array of pressure, temperature, and flow sensors to observe the state of the hydraulic system. The inverter and the pressure and temperature sensors all communicated over the CANopen protocol, the flow sensors gave analog inputs, and the safety valve was actuated with a digital signal. The CAN communication was interfaced by a Vector CAN/USB dongle~\cite{VectorCanUsbDongle}. All other IO went through a National Instruments data acquisition device (DAQ). In the new system, the forced change for the inverter to communicate over EtherCAT. This meant new IO hardware was needed to support this protocol, which was selected to be Beckhoff EtherCAT terminals. They provided the interface to connect the inverter, digital and analog IO, and a CAN bus for the sensor network~\cite{BeckhoffEL6751}. This made the DAQ unnecessary, but the CAN/USB dongle was still required due to deficiencies in Beckhoff's control interfaces, which are discussed with the control layer.

The system had been controlled by a conventional Windows workstation. This may have been fine to keep for the new system, but there was concern with using a whole workstation for a single experiment. Instead, one of Beckhoff's industrial PCs was selected as the computer to leverage Beckhoff's support~\cite{BeckhoffC60XX}. It was expensive, but it was thought that architecting the system to align with vendor expectations would make development easier and mitigate potential problems. However, the workstation did tap into the CAN bus to initialize the sensors, so it wasn't completely out of the loop.

In the old system, the workstation ran Windows, the DAQ was managed by drivers from National Instruments, and the CAN bus was initialized using Vector's CANalyzer application and runtime commands came through Vector's Simulink library~\cite{VectorCanalyzer, SimulinkVectorCAN}. Due to issues getting reliable EtherCAT connections with Simulink (possibly for user error, though the documentation wasn't simple), the new system uses a PLC-based middleware. Still based on Windows, it has libraries and runtimes to run the EtherCAT bus, to read from the CAN bus, and to interface with an external development computer. CANalyzer remained to initialize the CAN bus.

The old system had controllers written with Matlab and Simulink. However, due to middleware issues, they were abandoned in favor of writing controllers in PLC languages based on IEC~61131-3 with the TwinCAT development environment~\cite{IEC61131-3, BeckhoffTwinCAT}. The major issue with TwinCAT is that its CANopen implementation had no documented way to initialize devices. This was worked around by initializing devices with CANalyzer on the workstation.

Together, the stack for the old system was CANopen-based IO with a DAQ supplementing simple analog and digital signals, a workstation computer, Windows middleware with drivers and libraries to support communication with the CAN bus and DAQ, and a controller based on Matlab and Simulink. In the new system, the stack was a mix of EtherCAT and CANopen IO, a Beckhoff Industrial PC in concert with a workstation, PLC middleware with an application to initialize the CAN network, and a controller written in PLC languages with TwinCAT. This is illustrated in figure~\ref{fig:mechatronic_tech_stack_dolores}.
